Replication on multicore considering hierarchy and characteristics of the network

### 2013-03-20 paper.tex
Added a section on the network model

% tikz stuff
\usepackage{tikz}
\usepackage{pgfplots}
\usetikzlibrary{shapes,positioning,calc,snakes,arrows,shapes}

@@

\begin{document}

@@

\paragraph{What do you want to enable?} Increase algorithm performance
on multicores by applying distributed principles, particularly
replication. The challenge then is to guarantee consistency. Previous
work was using primary copy protocols to do that~\cite{Wiesmann2000}, a
passive replication scheme (i.e.\ nodes do not have to contact all
replicas). They do that because they claim that it is not suitable for
clients to know about replication details. An atomic broadcast is
often used as foundation of such algorithms. It guarantees in reliable
delivery of messages to every node.

\paragraph{What problem are you solving, and why is it hard?} Three problems:
\begin{description}
\item[placement] machine dependent
\item[consistency] need to consider machine \emph{characteristics};
  multicores are \emph{different} from classical distributed systems.
\item[machine characteristics] diverse, fast changing, complex. One
  example is the batch size for sending a broadcast.
\end{description}

\paragraph{What's the related work?} People do replication to deal
with scalability challenges on multicores, e.g.\
databases~\cite{Salomie2011, Wiesmann2000} and operating
systems~\cite{fos:osr09, tornado:osdi99, barrelfish:sosp09}. But they
don't to it properly. They do not consider machine characteristics or
apply the wrong distributed technologies.

\paragraph{What new idea(s) will solve the problems?} We do a
hierarchical group communication, which implements an atomic
broadcast on multicore machines. It has message complexity of
$O(\frac{n}{2})$ and time complexity $O(log(n))$ as opposed to
algorithms from traditional distributed systems having $O(2)$ message
complexity and $O(n)$ time complexity. %
\\
{\footnotesize Example: sequential consistency on shared memory, do
  not really need to have primary backup.}

\paragraph{How will you go about it?} Replicas on NUMA domains, show
that overall throughput is higher. Then, use a quorum based approach for
consistency. Consistency by doing a hierarchical group communication
approach instead of sequential send from same nodes.

\paragraph{How will you know and show it works?} Database?
Microbenchmark? Compare to primary backup, which is typically
used. Measure time required to apply an update.

\paragraph{What is your hypothesis?} We can do much better if machine
characteristics are considered.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Mechanisms}

\subsubsection{Aggregation}

@@

Open questions: How does this work with several senders (everyone is
potentially a sender). Maybe something like in navigation
systems. Every node has an entry point, which forwards the
messages. Nodes are then clustered \ldots

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Implementation}

@@

Reducing the number of children also helps in select. It reduces the
number of ``channels'' to poll, and therefore the latency of detecting
messages. Instead of

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Experimental results}

\paragraph{Current numbers}

@@

\paragraph{Tree setup} \stefan{describe}

\paragraph{Measurements to do next} %
\begin{itemize}
\item Increase number of cores to talk to in every round. Use a
  separate waitsets, so that we don't have additional overhead from
  the bindings not used in each concrete measurement.
\item Play with batching. Incrementally increase the number of
  messages send in a batch. I assume this is machine specific, which
  means, that this is something programmers do not want to deal
  with. We measure this at dynamically at run-time and find the right
  batch size.
\end{itemize}

\paragraph{Minimal cost of sending a tree-broadcast} %
Plot~\ref{pgfplot:201303141819} shows the minimal cost of flooding a
sub-tree. The group communication is based on a binary tree. Core 0 is
the root, cores 1 and 2 are its children etc.

@@

\begin{figure}
  \caption{Plot}
  \label{pgfplot:201303141819}
  \begin{tikzpicture}
    \begin{axis}[
      xlabel=core id,
      ylabel={cost for subtree [cycles]}]
    \addplot[color=red,mark=x] coordinates {
      (0,9475.0)
      (1,7935.0)
      (2,7894.0)
      (3,5965.0)
      (4,5604.0)
      (5,5718.0)
      (6,5647.0)
      (7,4316.0)
      (8,2989.0)
      (9,3173.0)
      (10,3395.0)
      (11,3332.0)
      (12,3284.0)
      (13,3310.0)
      (14,3231.0)
      (15,2542.0)
      (16,586.0)
      (17,588.0)
      (18,576.0)
      (19,583.0)
      (20,536.0)
      (21,539.0)
      (22,533.0)
      (23,516.0)
      (24,546.0)
      (25,552.0)
      (26,541.0)
      (27,543.0)
      (28,537.0)
      (29,527.0)
      (30,533.0)
      (31,534.0)
    };
    \end{axis}
  \end{tikzpicture}

@@

\begin{figure}
  \input{qrm_tree_gruyere}
  \caption{Tree of cores for broadcast on gruyere}
\end{figure}

@@

\begin{itemize}
\item Move server to a core != 0
\item Use low-level UMP stuff
{
\renewcommand{\labelitemi}{\checkmark}
\item disable yielding set 
}
\item some other item
\item run several request in parallel, give them IDs, can use in the
  measurement struct. Good for everything having transactions (DB,
  transactional memory, consistency in replication systems)
\end{itemize}

>>>>

% tikz stuff
\usepackage{tikz}
\usepackage{pgfplots}
% configuration
\usetikzlibrary{shapes,positioning,calc,snakes,arrows,shapes}
\pgfplotsset{width=.9\linewidth}

@@

% ==================================================
\setcounter{section}{0} % Start sections with 1, not 0
\begin{document}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

\paragraph{What do you want to enable?} Increase algorithm performance
on multicores by applying distributed principles to overcome
scalability challenges on such machines.

\paragraph{What problem are you solving, and why is it hard?} We are
applying distributed algorithms to multicores to enable better
scalability. However, characteristics of a multicore a different to
those of traditional distributed systems. Traditional algorithms need
to be reevaluated for multicores.

\paragraph{What's the related work?} People apply techniques from
distributed systems to deal with scalability challenges on multicores,
e.g.\ databases~\cite{Salomie2011, Wiesmann2000} and operating
systems~\cite{fos:osr09, tornado:osdi99, barrelfish:sosp09}. But they
don't to it properly. They do not consider machine characteristics or
apply the wrong distributed technologies. Also, what they do is rather
ad hoc.

\paragraph{What new idea(s) will solve the problems?} Instead of just
reusing distributed algorithms, we reevaluate them for multicores and
investigate the impact of the multicore network model on algorithm
performance.

\paragraph{How will you go about it?} We investigate the problem for
atomic-broadcasts since many problems in distributed systems can be
reduced to this communication primitive. Traditional implementations
are sending requests to all nodes in parallel to hide high propagation
delays. %
However, this does not work on multicores, since the propagation time
is not dominating any longer. %
Instead, we build a tree for communication to parallelize transmit and
receive cost across nodes. This has been done before (BF SOSP, IBM,
.. ), but not for the general case of multicore communication.

\paragraph{How will you know and show it works?} We will use a
microbenchmark measuring broadcast cost and compare our tree
implementation to the sequential case. Maybe we put some high-level
applications on top of it.

\paragraph{What is your hypothesis?} Performance of distributed
algorithms can be improved if characteristics of multicores are
considered when building and/or selecting them. %
For the broadcast example, this means: the time complexity for a
networks with dominating transmit and receive time will be $O(log(n))$
instead of $O(n)$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Related work}

A multicore aware broadcast has been implemented in~\cite{Tu2008}. It
takes the cache architecture of the machine into consideration for
building the broadcast. It is however for larger messages (starting at
4K) and it does not seem to be performing very well (numbers are given
in microseconds -6). 4K messages are 52 microseconds, which is 150000
cycles. For synchronization primitives, a cache line is sufficient.

Another tree implementation with MPI in mind was done
in~\cite{Graham2008}. They have comparable numbers to what we found.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Model}

The selection of distributed algorithms depends on the characteristics
of a network. Multicores networks are heterogeneous combinations of
complex networks. %
Their diversity makes manual selection tedious. Hence, we aim for an
automatic approach for selecting algorithms for a specific
network. %
We express the network characteristics in a multicore network
model. This model is currently by no means complete. We are going to
extend it over time to address more complexities of current and future
multicores.

The network model is defined by communication channels, which in turn
are defined by a set of attributes, i.e.\
\begin{itemize}
\item reliability (loss, link failure, in-order delivery, phantoms,
  corruption, duplicated messages): Ethernet is not reliable,
  multicore interconnect networks often are
\item synchronized clocks: sychronized clocks + bound on propagation
  time $\rightarrow$ \emph{synchronous network} model. \\Then:
  $\exists$ algorithm for synchronous networks $\rightarrow
  \exists$ algorithm for asynchronous
\item Security \& and failure detection (and the cost for that)
\end{itemize}

These are more about performance:
\begin{itemize}
\item ``breakdown'' of message send cost: Ethernet has dominating
  propagation time vs.\ dominating send and receive time on multicores
\item latency between each pair of nodes
\end{itemize}

% Model given
We assume that the model is given (if it is not, we can do online
measurements and read some specs to find the relevant information). We
just define one ourselves for now or build one manually for a
particular machine. In a real-world scenario, this does not work. The
model needs to be created automatically for a particular machine (due
to the diversity of machines, and also the pace at which hardware changes)

% Express as:
We express the model in a graph, where nodes are processors and edges
are (on-hop?) communication links.

% How to deduct tree from model?
We then use this graph to build a group-communication
tree. Interesting techniques could be: clustering, to find closely
connected subsets, distributed messages across cluster first, and then
within cluster. Maybe also a spanning-tree protocol, to find a minimal
graph such that all nodes can be reached.
% Tree -> scheduling
Assuming the tree, there is a scheduling problems for the order in
which to send messages to children. The intuition is to send on
long-delay links first to hide the more expensive latencies.

% Limitations
What we do not consider yet is link congestion.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Mechanisms}

\subsection{Aggregation}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Implementation}

We implement our quorum protocol on Barrelfish using the UMP
interconnect driver. UMP provides channels between
endpoints. Communication is reliable, and messages are delivered
in-order within a channel. Although UMP channels are uni-directional,
Barrelfish automatically established two uni-directional channels on
bind time. Hence, we consider them as being bi-directional. Every UMP
channels has independent buffers associated with it.

First, we establish connections such that we get a fully connected
graph with vertices being the quorum instances and edges being UMP
channels between them. For that, we need $O(n^2)$ UMP channels.

We build up UMP channels in rounds $r_i$. The quorum instance on core
$i$ is creating a UMP channel to all nodes $j$ with $j>i$ in
$r_i$. The beginning of a round needs to be synchronized such that
node $i+1$ only starts establishing connections when all other nodes
completed $i$. Synchronization is required since UMP channels are
scheduled round-robin. Messages on different channels are received
non-deterministically.

@@

We show an example broadcast tree for a 8x4x1 AMD Barcelona machine
(gruyere) in Figure~\ref{fig:qrm_tree_gruyere}.

\begin{figure}
  \input{qrm_tree_gruyere}
  \caption{Tree of cores for broadcast on gruyere}
  \label{fig:qrm_tree_gruyere}
\end{figure}

\paragraph{Observations} Reducing the number of children also helps in
select. It reduces the number of ``channels'' to poll, and therefore
the latency of detecting messages. Instead of

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Experimental results}

\paragraph{Current numbers} Current numbers are listed in
Table~\ref{tab:bc_measurements}.

@@

\paragraph{Measurements to do next} %
\begin{itemize}
\item Play with batching. Incrementally increase the number of
  messages send in a batch. I assume this is machine specific, which
  means, that this is something programmers do not want to deal
  with. We measure this at dynamically at run-time and find the right
  batch size.
\end{itemize}

\paragraph{Minimal cost of sending a tree-broadcast} %
Plot~\ref{pgfplot:201303141819} shows the cost of flooding a sub-tree
(with the 50\% worst measurements dropped). The group communication is
based on a binary tree. Core 0 is the root, cores 1 and 2 are its
children etc.

@@

\begin{figure}
  \caption{Execution time for a broadcast with ACK on gruyere. The
    cost is for execution for the sub-graph starting at the given
    node. Refer to Figure~\ref{fig:qrm_tree_gruyere} for a
    visualization of the broadcast tree used. }
  \label{pgfplot:201303141819}
  \begin{tikzpicture}
    \begin{axis}[
      xlabel=core id,
      scaled y ticks = false, % prevent 10^x stuff
      y tick label style={/pgf/number format/fixed},
      ylabel={cost for subtree [cycles]}]
    \addplot[
      color=red,
      very thin,
      mark=*,
      mark options={%
        scale=.4
      },
      error bars/y dir=both,
      error bars/y explicit] coordinates {
      (0,9932.3) +- (564.4,564.4)
      (1,8240.1) +- (298.5,298.5)
      (2,8259.6) +- (549.4,549.4)
      (3,6206.3) +- (225.1,225.1)
      (4,5914.9) +- (399.2,399.2)
      (5,6033.5) +- (462.2,462.2)
      (6,6070.4) +- (434.3,434.3)
      (7,4530.4) +- (188.6,188.6)
      (8,3163.4) +- (246.1,246.1)
      (9,3403.0) +- (285.3,285.3)
      (10,3650.7) +- (336.5,336.5)
      (11,3559.0) +- (318.8,318.8)
      (12,3558.1) +- (284.8,284.8)
      (13,3567.5) +- (304.5,304.5)
      (14,3476.3) +- (290.2,290.2)
      (15,2711.7) +- (151.9,151.9)
      (16,614.0) +- (17.5,17.5)
      (17,606.2) +- (17.3,17.3)
      (18,636.2) +- (72.5,72.5)
      (19,641.3) +- (46.3,46.3)
      (20,583.8) +- (46.7,46.7)
      (21,577.8) +- (38.3,38.3)
      (22,561.8) +- (18.1,18.1)
      (23,603.5) +- (56.6,56.6)
      (24,589.0) +- (33.4,33.4)
      (25,582.9) +- (33.2,33.2)
      (26,571.7) +- (16.3,16.3)
      (27,581.6) +- (36.6,36.6)
      (28,564.0) +- (16.4,16.4)
      (29,564.0) +- (20.8,20.8)
      (30,551.6) +- (13.0,13.0)
      (31,596.2) +- (51.4,51.4)
    };
    \end{axis}
  \end{tikzpicture}

@@

@@

\begin{itemize}
\item Move server to a core != 0
\item Use low-level UMP stuff
{
\renewcommand{\labelitemi}{\checkmark}
\item disable yielding
}
\item run several request in parallel, give them IDs, can use in the
  measurement struct. Good for everything having transactions (DB,
  transactional memory, consistency in replication systems)
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Future work}

\begin{itemize}
\item Need higher-level applications, such as:
  \begin{itemize}
  \item octopus
    \begin{itemize}
    \item higher level application?
    \end{itemize}
  \item capability system? \stefan{Ask Simon how expensive revocation
      currently is}
  \item a database with replication and consistency maintenance
    \begin{itemize}
    \item SharedDB? (buffers between operators, but no synchronization
      except for access to buffers, which are multiple writers, one
      reader)
    \item Crescando (some kind of state machine replication)
    \end{itemize}
  \end{itemize}
\item Machine model for topology aware tree
  \begin{itemize}
  \item Talked to Pravin about this a bit
  \end{itemize}
\item MPI collectives (as in~\cite{Tu2008})
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Open questions}

\paragraph{How does this work with several senders} 
Maybe something like in navigation systems. Every node has an entry
point, which forwards the messages. Nodes are then clustered
\ldots. Or: send the request up to the root and then down again? This
keeps the message complexity the same and doubles the time
complexity.
